---
title: "Analysis"
output:
  pdf_document: default
  word_document: default
date: "2024-04-15"
---

```{r setup, include=FALSE}
library(tidyverse)
library(readxl)
library(rvest)
library(scraEP)
library(arm)
library(patchwork)
library(pROC)
library(caret)
library(ROCR)
library(tinytex)

options(scipen = 999)
```

# Datascraping

## The Plan


## Transfers

### Data Collection

Scraping from transfermarkt

```{r,echo=FALSE}
window = c("s","w")
final_df = data.frame(matrix(nrow = 0, ncol = 7))
names(final_df) = c("In","Age","Position","Market_Value","Left","Fee","Year")
```

```{r,echo=FALSE}
for (i in 1:16) {
  for (j in 1:2){
    df = paste0("https://www.transfermarkt.com/premier-league/transfers/wettbewerb/GB1/plus/?saison_id=",2006+i,"&s_w=",window[j],"&leihe=0&intern=0&intern=1") %>%
      read_html() %>%
      html_nodes("table") %>%
      html_table(fill = T) %>%
      lapply(.,function(x) setNames(x,c("In","Age","Nat","Position","Market Value","Left","Fee"))) #load the ith year of tranfers
  
    df = df[-1] #remove the first tibble (doesnt feature transfers)
  
    year_df_i = as.data.frame(df[1]) %>% #
      mutate(Year = 2006+i) %>%
      mutate(Window = window[j])
    names(year_df_i) = c("In","Age","Position","Market_Value","Left","Fee","Year","Window")  #rename columns to match up
  
    for (k in 1:19) { #iterating through each other teams transfers in
      club_df = as.data.frame(df[2*k+1]) %>%
        mutate(Year = 2006+i)%>%
        mutate(Window = window[j])
      names(club_df)=c("In", "Age", "Position", "Market_Value", "Left", "Fee","Year","Window")
      year_df_i = rbind(year_df_i,club_df) #placing them at the end of the yearly df
  }
    final_df = rbind(final_df,year_df_i) #placing them at the end of the overall df
  }
}
final_df = final_df[,-3]
final_df = final_df[,-6]
final_df = final_df[,-4]

names(final_df) = c("Player","Age","Position","Market_Value","Left","Fee","Season","Window")
```

### Cleaning

Removing euro characters
```{r,echo=FALSE}
final_df$Fee = gsub("\\€|","",final_df$Fee)
final_df$Market_Value = gsub("\\€|","",final_df$Market_Value)
head(final_df)
```
Cleaning player names

```{r,echo=FALSE}
for(k in 1:length(final_df$Player)) {
  if (grepl(".",final_df$Player[k],fixed = T)==TRUE){  #if the player goes by 2 names (eg Brad Cooper)
    final_df$Player[k] = gsub("\\..*","",final_df$Player[k]) #remove everything after the '.'
    final_df$Player[k] = substr(final_df$Player[k],1,nchar(final_df$Player[k])-1) #and the character before '.'
  } else {  #if the player goes by 1 name (eg Fred)
    final_df$Player[k] = substr(final_df$Player[k],1,nchar(final_df$Player[k])/2) #remove half of the string
  }
}
final_df$Player = unaccent(final_df$Player)
head(final_df)
```

Cleaning fee and market value

```{r,echo=FALSE, warning=FALSE}
booltabmv = str_detect(final_df$Market_Value,"k")
booltabf = str_detect(final_df$Fee,"k")
final_df$Market_Value = gsub("\\m|","",final_df$Market_Value)
final_df$Fee = gsub("\\m|","",final_df$Fee)
final_df$Market_Value = gsub("\\k|","",final_df$Market_Value)
final_df$Fee = gsub("\\k|","",final_df$Fee)
final_df$Fee = as.numeric(final_df$Fee)
final_df$Market_Value = as.numeric(final_df$Market_Value)
final_df$Fee[is.na(final_df$Fee)] = 0
final_df$Market_Value[is.na(final_df$Market_Value)] = 0
final_df$Market_Value = final_df$Market_Value/1000^booltabmv
final_df$Fee = final_df$Fee/1000^booltabf
```

Removing 'no new arrivals'

```{r,echo = FALSE}
final_df %>%
  filter(Age == "No new arrivals") %>%
  head()
```


```{r,echo=FALSE}
final_df = subset(final_df, final_df$Age != "No new arrivals")
final_df %>%
  filter(Age == "No new arrivals")
```

Writing to CSV

```{r,echo=FALSE}
write.csv(final_df, "D:\\transfersall.csv",row.names = FALSE)
```




## Player Performance Statistics

### Selecting Clubs and Seasons

creating toread from presence

```{r,echo=FALSE}
presence = read_excel("D:\\LeaguePresence.xlsx")[,-1] #read in binary table of if team was in league in given year
clubs = names(presence) #the list of clubs that have been in the prem

toread = data.frame(matrix(ncol = 2, nrow = 0)) #initialising df for which teams to scrape
names(toread) = c("Club","SeasonStart")
```

```{r,echo=FALSE}
for (teams in 1:length(clubs)){  #iterating through teams
  for (seasonst in 1:15){ #iterating through seasons
    if (presence[seasonst,teams] == 1 & presence[seasonst+1,teams] == 1) {  #if they were in 2 seasons running
      check = as.data.frame(matrix(nrow = 0,ncol = 2))
      names(check) = c("Club","SeasonStart")
      check[1,] = c(clubs[teams],paste0(2006+seasonst,"-",2007+seasonst))
      check[2,] = c(clubs[teams],paste0(2007+seasonst,"-",2008+seasonst))
      toread = rbind(check,toread)  #add them and the season to the df
    }
  }
}
toread = toread[!duplicated(toread),]
head(toread)
```

### Data Collection

Scraping using toread

```{r,echo=FALSE}
total = as.data.frame(matrix(nrow = 0, ncol = 12))  #initialising df for playerbase
for (i in 1:length(toread$Club)) {  #iterating through the list of club-seasons
  starter = paste0("https://www.worldfootball.net/team_performance/",toread[i,1],"/eng-premier-league-",toread[i,2],"/") %>%
    read_html() %>%
    html_nodes("table") %>%
    html_table(fill = T) %>%
    lapply(.,function(x) setNames(x,c("Player", "Minutes", "Apps","Starts","Subbed_On",
                                      "Subbed off","Goals","Yellows","Second_Yellows","Reds")))
  #scraping the table for the team in the season
  df = as.data.frame(starter[2])[-1,]%>%
    mutate(Club = toread[i,1])%>%
    mutate(Season = toread[i,2])  #adding the club and season they were in
  total = rbind(total,df) #adding these players to the overall df
}
```

### Cleaning


```{r,echo=FALSE}
head(total)
```
Cleaning


```{r,echo=FALSE,warning=FALSE}
total$Minutes = gsub("\\'|","",total$Minutes)
total$Minutes[is.na(as.double(total$Minutes))]=0
total$Reds[is.na(as.double(total$Reds))]=0
total$Starts[is.na(as.double(total$Starts))]=0
total$Subbed_On[is.na(as.double(total$Subbed_On))]=0
total$Subbed.off[is.na(as.double(total$Subbed.off))]=0
total$Goals[is.na(as.double(total$Goals))]=0
total$Yellows[is.na(as.double(total$Yellows))]=0
total$Second_Yellows[is.na(as.double(total$Second_Yellows))]=0
total$Apps[is.na(as.double(total$Apps))]=0
total$Season = as.double(gsub("\\-.*","",total$Season))
total$Player = unaccent(total$Player)
total = total %>%
  mutate_at(c('Minutes','Apps','Starts','Subbed_On','Subbed.off','Goals','Yellows','Second_Yellows','Reds'),as.numeric)
head(total)
```

Write to csv

```{r,echo=FALSE}
write.csv(total, "D:\\appearance_stats.csv",row.names = FALSE)
```

## Team League Performance

### Data Collection

Scraping from skysports

```{r,echo=FALSE}
tot = as.data.frame(matrix(nrow = 0, ncol = 11))
names(tot) = c("Pos","Club","Pl","W","D","L","F","A","GD","Pts","Season")
clubs.new = c("Arsenal","Aston Villa","Birmingham","Blackburn",
              "Blackpool","Bolton","Bournemouth","Brentford",
              "Brighton","Burnley","Cardiff","Chelsea",
              "Crystal Palace","Everton","Fulham","Huddersfield",
              "Hull City","Leeds","Leicester","Liverpool",
              "Man City","Man Utd","Middlesbrough","Newcastle",
              "Norwich","Nottm Forest","Portsmouth","QPR",
              "Reading","Sheff Utd","Southampton","Stoke City",
              "Sunderland","Swansea","Tottenham","Watford",
              "West Brom","West Ham","Wigan","Wolves")
for(i in 1:16){
  starter = paste0("https://www.skysports.com/premier-league-table/",2006+i,"") %>%
    read_html() %>%
    html_nodes("table") %>%
    html_table(fill = T) %>%
    lapply(.,function(x) setNames(x,c("Pos","Club","Pl","W","D","L","F","A","GD","Pts")))
  yeardf = as.data.frame(starter[1])
  yeardf = yeardf[,-11] %>%
    mutate(Season = 2006+i)
  tot = rbind(tot,yeardf)
}
```

### Cleaning

Cleaning * and club names

```{r,echo=FALSE}
head(tot)
```


```{r,echo=FALSE}
for(i in 1:length(tot$Club)){
  if (str_detect(tot$Club,"\\*")[i] == TRUE){
    tot$Club[i] = gsub("\\*.*","",tot$Club[i])
    tot$Club[i] = substr(tot$Club[i],1,nchar(tot$Club[i])-1)}
}
```

```{r,echo=FALSE}
renaming = tot %>%
  group_by(Club) %>%
  summarise(n=n()) %>%
  arrange(Club)
oldnames = renaming$Club
oldnames = setdiff(oldnames,c("Derby County"))

for(i in 1:length(tot$Club)){
  for(j in 1:length(clubs.new)){
    if(tot$Club[i] == oldnames[j]){
      tot$Club[i] = clubs.new[j]
    }
  }
}
```

```{r,echo = FALSE}
head(tot)
```

Write to csv

```{r,echo=FALSE}
write.csv(tot, "D:\\league_tables.csv",row.names = FALSE)
```

# Analysis

## Building the Full Dataframe

```{r,echo=FALSE}
transfers = read.csv("D:\\transfersall.csv")
playerstats = read.csv("D:\\appearance_stats.csv")
teamstats = read.csv("D:\\league_tables.csv")
```

```{r,echo=FALSE}
test = inner_join(transfers, playerstats, by = c("Player","Season"), relationship = "many-to-many")
```

Cleaning duped values

```{r,echo=FALSE}
test %>%
  filter(Player == "Dimitar Berbatov") %>%
  filter(Season == 2008)
```

```{r,echo = FALSE}
presence = read_excel("D:\\LeaguePresence.xlsx")[,-1] #read in binary table of if team was in league in given year
clubs = names(presence) #the list of clubs that have been in the prem

for(i in 1:40){
  test$Club[test$Club == clubs[i]] = clubs.new[i]
}
```


```{r,echo=FALSE}
fulldf = as.data.frame(matrix(nrow = 0, ncol = 18))
names(fulldf) = names(test)
for(i in 1:length(test$Player)){
  if(test$Club[i] != test$Left[i]){
    fulldf = rbind(fulldf, test[i,])
  }
}
```

Adding results frame

```{r,echo=FALSE}
fulldf = inner_join(fulldf,teamstats, by = c("Club", "Season"),relationship = "many-to-many")
```

## Exploratory Analysis

### Change in Points per Game

Computing ppg diff

```{r,echo=FALSE}
fulldf = fulldf %>%
  mutate(LastSeasonPts=0,LastSeasonGD=0)
foo = fulldf %>%
  group_by(Club,Season) %>%
  summarise(Pts = mean(Pts),GD = mean(GD))
head(foo)
```

```{r,echo=FALSE}
for(entry in 1:length(fulldf$Player)){
  if(length(fulldf[fulldf$Club == fulldf$Club[entry] & fulldf$Season == fulldf$Season[entry]-1,]!=0)){
    fulldf$LastSeasonPts[entry] = foo$Pts[foo$Club == fulldf$Club[entry] & foo$Season == fulldf$Season[entry]-1]
    fulldf$LastSeasonGD[entry] = foo$GD[foo$Club == fulldf$Club[entry] & foo$Season == fulldf$Season[entry]-1]
  }
}
```

Removing 1st season

```{r,echo=FALSE}
fulldf = setdiff(fulldf, fulldf[fulldf$Season == "2007",])
```

```{r,echo=FALSE}
fulldf = fulldf %>%
  mutate(PPG.Before = LastSeasonPts/38, PPG.After = Pts/38) %>%
  mutate(PPG.Diff = PPG.After - PPG.Before) %>%
  mutate(GD.Change = GD - LastSeasonGD)
```

Histograms of PPG Diff before and after removing promoted teams

```{r,echo=FALSE}
hist(fulldf$PPG.Diff,breaks = 20, prob = TRUE, main = "Distribution of PPG.Diff",xlab = "PPG.Diff")
curve(dnorm(x,mean = mean(fulldf$PPG.Diff),sd = sd(fulldf$PPG.Diff)),add = TRUE)
```


```{r,echo=FALSE}
notpromoted = fulldf %>%
  filter(LastSeasonPts != 0)

hist(notpromoted$PPG.Diff, breaks = 20, prob = T, xlab = "PPG Difference", main = "Distribution of PPG Diff")
mean(notpromoted$PPG.Diff)
sd(notpromoted$PPG.Diff)
curve(dnorm(x,mean = mean(notpromoted$PPG.Diff),sd = sd(notpromoted$PPG.Diff)),add = TRUE)
```

### Fee

Investigating fee

```{r,echo=FALSE}
hist(notpromoted$Fee, breaks = 20, prob = T, xlab = "Fee", main = "Distribution of Fee")
alpha.est = (mean(notpromoted$Fee)^2)/var(notpromoted$Fee)
beta.est = (mean(notpromoted$Fee))/var(notpromoted$Fee)
curve(dgamma(x,alpha.est,beta.est), add = TRUE)
```

Position-wise fee

```{r,echo=FALSE}
ggplot(notpromoted, aes(y = log(Fee))) + geom_boxplot() + geom_hline(yintercept = log(mean(notpromoted$Fee)),col = "blue")+ facet_wrap(~Position)
```

```{r, include = FALSE}
notpromoted$Minutes = as.integer(notpromoted$Minutes)
notpromoted$Market_Value = as.double(notpromoted$Market_Value)
notpromoted$Season = as.integer(notpromoted$Season)
notpromoted$Apps = as.integer(notpromoted$Apps)
```

Fee against other variables

```{r,echo=FALSE}
plot1 = ggplot(notpromoted, aes(x = Minutes, y = Fee)) + geom_smooth()
plot2 = ggplot(notpromoted, aes(x = Apps, y = Fee)) + geom_smooth()
plot3 = ggplot(notpromoted, aes(x = Season, y = Fee)) + geom_smooth()
plot4 = ggplot(notpromoted, aes(x = Pts, y = Fee)) + geom_smooth()
plot1+plot2+plot3+plot4
```

### Club League Standing

League standing fee-wise

```{r,echo=FALSE}
ggplot(fulldf, aes(x = Pos, y = Fee)) + geom_smooth()
```

Confidence intervals for top team vs bottom team

```{r,echo=FALSE}
first = fulldf %>%
  filter(Pos == 1)
bar = mean(first$Fee)

n = length(first$Fee)
xbar = mean(first$Fee)
s = sd(first$Fee)

margin= qt(0.975, n-1)*s/sqrt(n)

low = xbar - margin
up = xbar + margin

firstconfint = c(low,xbar,up)

last = fulldf %>%
  filter(Pos == 20)
bar = mean(last$Fee)

n = length(last$Fee)
xbar = mean(last$Fee)
s = sd(last$Fee)

margin = qt(0.975, n-1)*s/sqrt(n)
low = xbar - margin
up = xbar + margin

lastconfint = c(low,xbar,up)

firstconfint
lastconfint
```

### Issue with PPG.Diff

Low effect of ppg.diff on fee

```{r,echo=FALSE,warning=FALSE}
ggplot(notpromoted[notpromoted$Fee > 1,], aes(x = PPG.Diff, y = log(Fee))) + geom_point() + geom_smooth()
```

# Fitting a Model

## Defining Success

Success as whether media deems you successful

```{r,echo=FALSE}
successes = c(c("Diego Costa",2014),c("Fernando Torres",2007),c("Sergio Aguero",2011),c("Michu",2012),
              c("N Golo Kante",2015),c("Erling Haaland",2022),c("Mo Salah",2017),c("Ayoze Pérez",2014),
              c("Harry Maguire",2017),c("Romelu Lukaku",2014),c("Wilfried Zaha",2014),c("Philippe Coutinho",2012),
              c("Dele Alli",2015),c("Kevin De Bruyne",2015),c("Sadio Mane",2014),c("Cesar Azpilicueta",2012),
              c("David de Gea",2011),c("Yaya Toure",2010),c("Eden Hazard",2012),c("Luis Suarez",2010),
              c("Robin Van Persie",2012),c("Andy Robertson",2017),c("Virgil van Dijk",2017),c("David Silva",2010),
              c("Vincent Kompany",2008),c("Pablo Zabaleta",2008),c("Fernandinho",2013),c("Raheem Sterling",2016),
              c("Luka Modric",2008),c("Mesut Ozil",2013),c("Nemanja Matic",2013),c("Bruno Fernandez",2019),
              c("Cesc Febregas",2014),c("Oscar",2012),c("Hugo Lloris",2012),c("Alisson",2018),
              c("Ederson",2017),c("Rodri",2019),c("Ruben Dias",2020),c("Zlatan Ibrahimovic",2016))

failures = c(c("Danny Drinkwater",2017),c("Dani Osvaldo",2013),c("Mario Balotelli",2014),c("Alexis Sanchez",2018),
             c("Bebe",2010),c("David Bentley",2008),c("Roger Johnson",2011),c("Alberto Aquilani",2009),
             c("Gaston Ramirez",2012),c("Jack Rodwell",2014),c("Saido Berahino",2017),c("Giannelli Imbula",2016),
             c("Nikola Zigic",2010),c("Papy Djilobodji",2016),c("Didier Ndong",2016),c("Andy Carroll",2011),
             c("Alvaro Morata",2017),c("Sebastien Haller",2019),c("Tanguy Ndombele",2019),c("Eliaquim Mangala",2014),
             c("Tiemoue Bakayoko",2017),c("Nicolas Pepe",2019),c("Lazar Markovic",2014),c("Shkodran Mustafi",2016),
             c("Angel Di Maria",2014),c("Paul Pogba",2016),c("Romelu Lukaku",2021),c("Harry Maguire",2019))
successes
failures
```

Creating booking points and success

```{r,echo=FALSE}
fulldf$Reds = as.numeric(fulldf$Reds)
fulldf$Yellows = as.numeric(fulldf$Yellows)
fulldf = fulldf %>%
  mutate(Booking_Pts = Yellows + 2*Reds)
training = fulldf %>%
  mutate(Success = -1)
for(i in 1:length(successes)/2){
  training$Success[training$Player == successes[2*i-1] & training$Season == successes[2*i]] = 1
}
for(i in 1:length(failures)/2){
  training$Success[training$Player == failures[2*i-1] & training$Season == failures[2*i]] = 0
}
training[,9:17] = sapply(training[,9:17],as.numeric)
```

Creating validation set

```{r,echo=FALSE}
trainingdf = training %>%
  filter(Success != -1)
test = setdiff(training,trainingdf)
```

## An Initial Model

### Mathematical Background



### The Model

First model

```{r,echo=FALSE}
mod0 = glm(data = trainingdf, formula = Success ~ I(Minutes/45) + LastSeasonPts:PPG.Diff + Apps:Yellows + Apps:Reds + Pos:Fee, family = binomial(logit))
summary(mod0)
```

## Improving on the Model

Adjustment for improvements

```{r,echo=FALSE}
mod0 = glm(data = trainingdf, formula = Success ~ I(Minutes/50) + Pts + Apps:Booking_Pts + Season:Fee , family = binomial(logit))
summary(mod0)
```

Creating validation set for a new model

```{r,echo=FALSE}
training = fulldf %>%
  mutate(Success = -1)
  

for(i in 1:length(successes)/2){
  training$Success[training$Player == successes[2*i-1] & training$Season == successes[2*i]] = 1
}
for(i in 1:length(failures)/2){
  training$Success[training$Player == failures[2*i-1] & training$Season == failures[2*i]] = 0
}

training[,9:17] = sapply(training[,9:17],as.numeric)

training = training %>%
  filter(Success != -1)

```

New model based on observations of old models

```{r,echo = FALSE}
set.seed(442)

trainingdf = training[sample(nrow(training),size = 45),]

test = setdiff(training,trainingdf)

mod1 = glm(data = trainingdf, Success ~ Window:Minutes + Pts + Season:Fee + Booking_Pts,family = binomial(logit))
summary(mod1)
```

Position-wise bookings

```{r,echo = FALSE}
trainingdf %>%
  group_by(Position) %>%
  summarise(Mean.Booking_Pts = mean(Booking_Pts), n = n())
```

Computing Minutes_Value and fitting final model

```{r,echo=FALSE}
notpromoted = fulldf %>%
  filter(LastSeasonPts != 0)
winter = notpromoted %>%
  filter(Window == as.factor("w")) %>%
  mutate(Minutes_Value = 2*Minutes)
summer = notpromoted %>%
  filter(Window == as.factor("s")) %>%
  mutate(Minutes_Value = Minutes)
notpromoted = rbind(winter,summer)

training = notpromoted %>%
  mutate(Success = -1)
  

for(i in 1:length(successes)/2){
  training$Success[training$Player == successes[2*i-1] & training$Season == successes[2*i]] = 1
}
for(i in 1:length(failures)/2){
  training$Success[training$Player == failures[2*i-1] & training$Season == failures[2*i]] = 0
}

training[,9:17] = sapply(training[,9:17],as.numeric)

training = training %>%
  filter(Success != -1)

set.seed(442)

trainingdf = training[sample(nrow(training),size = 45),]

test = setdiff(training,trainingdf)

mod2 = glm(data = trainingdf, Success ~ Minutes_Value + Apps:Booking_Pts + LastSeasonPts:Pts + Season:Fee,family = binomial(logit))
summary(mod2)
```

## Validation

### Cross Reference

Checking validation of new data

```{r, echo = FALSE}
test = test %>%
  mutate(Prediction = predict(mod1, test, type = "response")) %>%
  dplyr::select(c(Player,Success,Prediction))
test
```

### Accuracy

Computing accuracy

```{r,echo=FALSE}
acc <- NULL
for(i in 1:nrow(training))
{
  train <- training[-i,]
  test <- training[i,]
  
  model <- mod1
  results_prob <- predict(model,test,type='response')
  
  results <- ifelse(results_prob > 0.5,1,0)
  
  answers <- test$Success
  
  wrong <- mean(answers != results)
  
  acc[i] <- 1-wrong
}
mean(acc)
```

First model accuracy

```{r,echo=FALSE}
acc <- NULL
for(i in 1:nrow(training))
{
  train <- training[-i,]
  test <- training[i,]
  model = mod0
  results_prob <- predict(model,test,type='response')
  
  results <- ifelse(results_prob > 0.5,1,0)
  
  answers <- test$Success
  
  wrong <- mean(answers != results)
  
  acc[i] <- 1-wrong
}

mean(acc)
```

Apply new model to full set of unpromoted teams

```{r,echo=FALSE}
notpromoted$Success = predict(mod2, notpromoted, type = "response")
```

## Checking Model Assumptions

Checking linearity of minutes vs Success

```{r,echo=FALSE}
ggplot(training, aes(x = Minutes_Value, y = Success)) + geom_smooth()
```

## Case Study - Leicester City FC

Checking Leicester's 2015 season

```{r,echo=FALSE}
Leicester = notpromoted %>%
  filter(Club == "Leicester") %>%
  filter(Season == 2015)
Leicester = Leicester[,append(1:17,36)]
Leicester = Leicester[,-c(3,5)]
Leicester
```

& 2016 season

```{r,echo=FALSE}
Leicesterbad = notpromoted %>%
  filter(Club == "Leicester") %>%
  filter(Season == 2016)
Leicesterbad = Leicesterbad[,append(1:17,36)]
Leicesterbad = Leicesterbad[,-c(3,5)]
Leicesterbad
```
